---
title: "Multiple Correspondence analysis of CBCL behavioral data"
output: html_document
params:
    cbcl_infile: "../../Synthetic_data/synthetic_cbcl_data.csv"
    cbcl_labels: "cbcl_labels.csv"
    cbcl_filtered_outfile: "cbcl_filtered.csv"
    cbcl_imputed_outfile: "cbcl_imputed.csv"
    cbcl_mca_coord_outfile: "cbcl_mca_coord.csv"
    cbcl_mca_outfile: "cbcl_mca.RData"
    imputation_method: 
        label: "Type of phenotype imputation method."
        value: "RF"
        input: select
        choices: ["RF", "MCA"]
    missing_value_cutoff:
        label: "Threshold for removing individuals with missing values."
        value: 0.5
author:
    - Anna Niehues^[Radboud University Medical Center, Anna.Niehues@radboudumc.nl]
---

This notebook performs filtering based on missingness, missing-value imputation, and multiple correspondence analysis (MCA). Visualizations of data and results are plotted.

```{r}
print(params)
```

Read phenotypic data (samples in rows, survey elements in columns).

```{r}
# read phenotypic data
pheno_in <- read.csv(params$cbcl_infile, row.names = 1)
vars_in <- colnames(pheno_in)
# labels 
pheno_labels <- read.csv(params$cbcl_labels)
knitr::kable(pheno_labels)
# convert 999999 to NA
pheno_in[pheno_in == "999999"] <- NA
# convert to factor
pheno_in[, vars_in] <- lapply(pheno_in[, vars_in], as.factor)
```

## Overview of input data

```{r}
summary(pheno_in)
# distribution of missing values across samples and features
# histogram of missing values per sample
hist(rowSums(is.na(pheno_in)))
# histogram of missing values per feature
hist(colSums(is.na(pheno_in)))
```

## Filter phenotypic data

Filter features: remove non-informative (invariant) features

```{r}
pheno_filtfeat <- pheno_in[, sapply(pheno_in, nlevels) > 1]
message("[FILTERING] Removed ", dim(pheno_in)[2]-dim(pheno_filtfeat)[2], 
        " invariant feature(s).")
```

Filter samples: remove all samples that have missing values across most features.

```{r}
pheno_filtsamp <- pheno_filtfeat[
  rowSums(is.na(pheno_filtfeat)) < params$missing_value_cutoff*dim(pheno_filtfeat)[2],]
message("[FILTERING] Removed ", dim(pheno_filtfeat)[1]-dim(pheno_filtsamp)[1], 
        " samples with more than ", params$missing_value_cutoff*100, 
        "% missing values.")
```

```{r}
# remove invariant features
pheno_filt <- pheno_filtsamp[, sapply(pheno_filtsamp, nlevels) > 1]
message("[FILTERING] Removed ", dim(pheno_filtsamp)[2]-dim(pheno_filt)[2], 
        " invariant feature(s).")
```

## Overview of filtered data

```{r}
message("[FILTERING] Filtered data set contains ", dim(pheno_filt)[1], 
        " samples, ", dim(pheno_filt)[2], " features, and ", 
        sum(is.na(pheno_filt)), " missing values.")
print(summary(pheno_filt))
# histogram of missing values per sample
hist(rowSums(is.na(pheno_filt)))
# histogram of missing values per feature
hist(colSums(is.na(pheno_filt)))
# write filtered data to output file
write.csv(pheno_filt, params$cbcl_filtered_outfile)
```

## Impute missing values

Random forests imputation

```{r}
#' Non-parametric missing value imputation in survey data using random forests
#' 
#' @param df Data frame containing subjects in rows and questions in columns 
impute_cbcl_rf <- function(df) {
  library(missForest)
  stopifnot(all(sapply(df, is.factor)))
  # hyperparameters: ntree, mtry, replace, sampsize, nodesize, maxnodes
  # suggestion: could also make use of additional phenotype information 
  # impute values
  df_imp <- missForest(
    df,
    ntree = 1000,
    variablewise = TRUE)
  return(df_imp)
}
```

Imputation using MCA

```{r}
#' Imputing missing categorical data using Multiple Correspondence Analysis
#'  
#' @param df Data frame containing subjects in rows and questions in columns
#' @return list of different plots of MCA
impute_cbcl_mca <- function(df){
  library(missMDA)
  stopifnot(all(sapply(df, is.factor)))
  # estimate the number of dimensions for MCA by cross-validation
  ncp_est <- estim_ncpMCA(
    df, 
    ncp.min = 0, 
    ncp.max = 7,
    method = "Regularized", 
    method.cv = "Kfold",
    verbose = TRUE) 
  print(ncp_est)
  # impute missing values using estimated number of dimensions
  df_imputed <- imputeMCA(df, ncp = ncp_est$ncp)
  return(df_imputed)
}
```

Perform imputation based using chosen method 

```{r}
if (params$imputation_method == "RF") {
  pheno_rfimp_res <- impute_cbcl_rf(pheno_filt)
  names(pheno_rfimp_res)
  # out-of-bag (OOB) imputation error estimate
  pheno_rfimp_res$OOBerror
  # extract imputed (complete) data
  pheno_imp <- pheno_rfimp_res$ximp
} else if (params$imputation_method == "MCA") {
  pheno_mcaimp_res <- impute_cbcl_mca(pheno_filt)
  names(pheno_mcaimp_res)
  # extract imputed (complete) data
  pheno_imp <- pheno_mcaimp_res$completeObs
  summary(pheno_imp)
}
summary(pheno_imp)
write.csv(pheno_imp, params$cbcl_imputed_outfile)
```

```{r}
# prepare data for plot
library(reshape2)
library(stringr)
tmp_pheno_list <- list(filtered = pheno_filt, imputed = pheno_imp)
tmp_pheno_list <- lapply(names(tmp_pheno_list), function(x) {
  df <- tmp_pheno_list[[x]]
  df$data <- x
  df
})
pheno_all_wide <- do.call(rbind, tmp_pheno_list)
pheno_all_long <- melt(pheno_all_wide, id.vars = c("data"))
pheno_labels$variable_names <- make.names(pheno_labels$variable)
pheno_all_long$label <- apply(pheno_all_long, 1, function(x) {
  label <- pheno_labels[
    pheno_labels$variable_names == x[["variable"]], ][["label"]]
  str_wrap(gsub("CBCL6_18 - ", "", label), width = 12)
})

rm(tmp_pheno_list)
rm(pheno_all_wide)

# plot distribution of values, before and after imputation
library(ggplot2)
cbf_palette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
lapply(split(names(pheno_filt), ceiling(seq_along(names(pheno_filt)) / 4)), 
       function(x) {
         ggplot(pheno_all_long[pheno_all_long$variable %in% x,]) +
           geom_bar(aes(x = value, fill = value)) +
           scale_fill_manual(values = cbf_palette) +
           facet_grid(label ~ data, scales = "free_y")
})
```

## Perform multiple correspondence analysis

Multiple correspondence analysis (MCA) is a dimension reduction technique that can be used for categorical data sets. The results are numerical variables (factors) representing the data. 

```{r}
#' Multiple Correspondence Analysis of survey question data with missing values
#'  
#' @param df Data frame containing subjects in rows and questions in columns
#' @return list of different plots of MCA
perform_mca <- function(df){
  library(FactoMineR)
  stopifnot(all(sapply(df, is.factor)))
  mca <- MCA(
    df, 
    ncp = 10, 
    graph = FALSE,
    method = "Indicator" # "Burt" 
  )
  return(mca)
}

names(pheno_imp) <- lapply(names(pheno_imp), function(x) {
  label <- pheno_labels[
    pheno_labels$variable_names == x, ][["label"]]
  str_wrap(gsub("CBCL6_18 - ", "", label), width = 16)
})

mca_pheno <- perform_mca(pheno_imp)
```

### Quality of dimension reduction

Scree plot - percentage of variation (inertia) explained by dimensions.

```{r}
library(factoextra)
knitr::kable(
  head(get_eigenvalue(mca_pheno), n = 10),
  caption = "Eigenvalues and explained variance for MCA dimensions 1-10.")
fviz_screeplot(mca_pheno, addlabels = TRUE) +
  geom_point()
```

### Relative contribution of individuals and variables 


Biplot of individuals and behavior variable categories

```{r}
fviz_mca_biplot(
  mca_pheno,
  repel = TRUE,
  label = "var",
  ggtheme = theme_minimal())
```

#### Correlation between **variables** and MCA dimensions

```{r}
fviz_mca_var(
  mca_pheno, 
  choice = "mca.cor", 
  axes = c(1, 2),
  repel = TRUE, 
  ggtheme = theme_minimal())
```

#### Coordinates of **variable categories** and contribution to dimensions

```{r}
fviz_mca_var(
  mca_pheno, 
  col.var = "contrib", # color by contribution to dimension
  repel = TRUE, 
  gradient.cols = c("#E69F00", "#56B4E9"),
  ggtheme = theme_minimal())
```

```{r}
lapply(1:10, function(dim) {
  # contribution to MCA dimensions
  fviz_contrib(mca_pheno, choice = "var", axes = dim, top = 10)
})
```


#### Degree of association of **variable categories** with dimensions

How well is a variable category represented by the dimensions? 

```{r}
fviz_mca_var(
  mca_pheno, 
  col.var = "cos2", # color by quality of representation
  repel = TRUE, 
  gradient.cols = c("#E69F00", "#56B4E9"),
  ggtheme = theme_minimal())
```

```{r}
lapply(1:10, function(dim) {
  # quality of representation (cos2), squared cosine
  fviz_cos2(mca_pheno, choice = "var", axes = dim, top = 10)
})
```


MCA dimension description

```{r}
dimdesc_mca <- dimdesc(mca_pheno, axes = 1:10)
dimdesc_mca
```

Write MCA results to file

```{r}
mca_coord <- get_mca_ind(mca_pheno)$coord
write.csv(mca_coord, params$cbcl_mca_coord_outfile)
save(mca_pheno, file = params$cbcl_mca_outfile)
```

```{r}
sessionInfo()
```

