---
title: "SNF downstream analysis with GEE"
author: 
  - Casper de Visser^[Radboud University Medical Center, casper.devisser@radboudumc.nl]
output: html_document
params:
  input_file_snf_pheno: snf_analysis_out.csv
---

### Load SNF dataframe
```{r}
print(params$input_file_snf_pheno)
df_snf_phenotypes <- read.csv(params$input_file_snf_pheno, row.names=1)
```


### GEE model to correct for twin effect
```{r, message = FALSE, warning = FALSE}
library(gee)


#' Function for GEE model
#'
#' @param data Dataframe with pheno metadata
#' @param pheno_feature Feature to select as outcome
#' @return GEE model
geemodel <- function (data, pheno_feature, snf_label) {

f <- as.formula(
	paste(pheno_feature,
		paste(snf_label),
		sep = " ~ "))

invisible(gee(f, data=data, id=Familynumber, family=gaussian, corstr="exchangeable", maxiter=100, na.action=na.omit, silent=T))
}



#' Function to run the GEE model with mofa factors as predictor for phenotypic feature
#'
#' @param data Dataframe with pheno metadata
#' @param pheno_feature Feature to select as outcome
#' @return end table with estimates and p-values
run_geemodel <- function(data, pheno_feature, snf_label) {


# Make dataframes for storing statistics
estimates <- matrix(NA, 10, 2)
robustSE <- matrix(NA, 10, 2)
robustZ <- matrix(NA, 10, 2)
pval <- matrix(NA, 10, 2)


coeff <- summary(invisible(geemodel(data, pheno_feature, snf_label)))$coefficients
coeff <- data.frame(coeff)
coeff$pval <- 2*pnorm(-abs(coeff[,5]))
coeff$padjusted <- p.adjust(coeff$pval)

return(coeff)

}

```

# GEE models with 4 SNF clusters as predictor

### Aggression score as outcome
```{r,  message = FALSE, warning = FALSE, results = 'hide'}
geemodel_agg <- run_geemodel(df_snf_phenotypes, 'Aggression_Tscore', 'SNF_label_4')
```

### Aggression score as outcome
```{r, echo=FALSE}
knitr::kable(geemodel_agg, caption="GEE model with Aggression T-score as outcome")
```


### MCA dimension coordinates as outcome
```{r, message = FALSE, warning = FALSE, results='hide'}
geemodel_mca1 <- run_geemodel(df_snf_phenotypes, 'Dim_1', 'SNF_label_4')
geemodel_mca2 <-run_geemodel(df_snf_phenotypes, 'Dim_2', 'SNF_label_4')
geemodel_mca3 <-run_geemodel(df_snf_phenotypes, 'Dim_3', 'SNF_label_4')
geemodel_mca4 <-run_geemodel(df_snf_phenotypes, 'Dim_4', 'SNF_label_4')
geemodel_mca5 <-run_geemodel(df_snf_phenotypes, 'Dim_5', 'SNF_label_4')
geemodel_mca6 <-run_geemodel(df_snf_phenotypes, 'Dim_6', 'SNF_label_4')
geemodel_mca7 <-run_geemodel(df_snf_phenotypes, 'Dim_7', 'SNF_label_4')
geemodel_mca8 <-run_geemodel(df_snf_phenotypes, 'Dim_8', 'SNF_label_4')
geemodel_mca9 <-run_geemodel(df_snf_phenotypes, 'Dim_9', 'SNF_label_4')
geemodel_mca10 <-run_geemodel(df_snf_phenotypes, 'Dim_10', 'SNF_label_4')
```

### Print GEE models
```{r, echo=FALSE}
knitr::kable(geemodel_mca1, caption="GEE model with MCA dimension 1 as outcome")
knitr::kable(geemodel_mca2, caption="GEE model with MCA dimension 2 as outcome")
knitr::kable(geemodel_mca3, caption="GEE model with MCA dimension 3 as outcome")
knitr::kable(geemodel_mca4, caption="GEE model with MCA dimension 4 as outcome")
knitr::kable(geemodel_mca5, caption="GEE model with MCA dimension 5 as outcome")
knitr::kable(geemodel_mca6, caption="GEE model with MCA dimension 6 as outcome")
knitr::kable(geemodel_mca7, caption="GEE model with MCA dimension 7 as outcome")
knitr::kable(geemodel_mca8, caption="GEE model with MCA dimension 8 as outcome")
knitr::kable(geemodel_mca9, caption="GEE model with MCA dimension 9 as outcome")
knitr::kable(geemodel_mca10, caption="GEE model with MCA dimension 10 as outcome")
```




# GEE models with 2 SNF clusters as predictor

### Aggression score as outcome
```{r,  message = FALSE, warning = FALSE, results = 'hide'}
geemodel_agg <- run_geemodel(df_snf_phenotypes, 'Aggression_Tscore', 'SNF_label_2')
```

### Aggression score as outcome
```{r, echo=FALSE}
knitr::kable(geemodel_agg, caption="GEE model with Aggression T-score as outcome")
```


### MCA dimension coordinates as outcome
```{r, message = FALSE, warning = FALSE, results='hide'}
geemodel_mca1 <- run_geemodel(df_snf_phenotypes, 'Dim_1', 'SNF_label_2')
geemodel_mca2 <-run_geemodel(df_snf_phenotypes, 'Dim_2', 'SNF_label_2')
geemodel_mca3 <-run_geemodel(df_snf_phenotypes, 'Dim_3', 'SNF_label_2')
geemodel_mca4 <-run_geemodel(df_snf_phenotypes, 'Dim_4', 'SNF_label_2')
geemodel_mca5 <-run_geemodel(df_snf_phenotypes, 'Dim_5', 'SNF_label_2')
geemodel_mca6 <-run_geemodel(df_snf_phenotypes, 'Dim_6', 'SNF_label_2')
geemodel_mca7 <-run_geemodel(df_snf_phenotypes, 'Dim_7', 'SNF_label_2')
geemodel_mca8 <-run_geemodel(df_snf_phenotypes, 'Dim_8', 'SNF_label_2')
geemodel_mca9 <-run_geemodel(df_snf_phenotypes, 'Dim_9', 'SNF_label_2')
geemodel_mca10 <-run_geemodel(df_snf_phenotypes, 'Dim_10', 'SNF_label_2')
```

### Print GEE models
```{r, echo=FALSE}
knitr::kable(geemodel_mca1, caption="GEE model with MCA dimension 1 as outcome")
knitr::kable(geemodel_mca2, caption="GEE model with MCA dimension 2 as outcome")
knitr::kable(geemodel_mca3, caption="GEE model with MCA dimension 3 as outcome")
knitr::kable(geemodel_mca4, caption="GEE model with MCA dimension 4 as outcome")
knitr::kable(geemodel_mca5, caption="GEE model with MCA dimension 5 as outcome")
knitr::kable(geemodel_mca6, caption="GEE model with MCA dimension 6 as outcome")
knitr::kable(geemodel_mca7, caption="GEE model with MCA dimension 7 as outcome")
knitr::kable(geemodel_mca8, caption="GEE model with MCA dimension 8 as outcome")
knitr::kable(geemodel_mca9, caption="GEE model with MCA dimension 9 as outcome")
knitr::kable(geemodel_mca10, caption="GEE model with MCA dimension 10 as outcome")
```


```{r}
sessionInfo()
```


